OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-02 main.tex
Wrote the first paragraph of the results.

\title[Dryhic]{OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes}

@@

\abstract{
  \textbf{Motivation:} The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented advance in our understanding of genome
structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but none of them addresses the common issue of
abnormal karyotypes. Many experiments are performed in cancer cell lines,
which typically harbor large-scale copy number variations that create
visible defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.\\
	\textbf{Results:}
We observed that current normalization methods perform badly in the
presence of large-scale copy number variations, obscuring biological
viarations and enhancing batch effects. To address this issue, we
developed an alternative appoach designed to take into account chromosomal
abnormalities. The method, called OneD, increases reproducibility among
replicates of Hi-C samples with abnormal karyotype, thus significantly
outperforming previous methods.  In the absence of chromosomal
aberrantions, OneD fared equally well as state-of-the-art methods, making
it a safe choice for Hi-C normalization. OneD is fast and scales well in
terms of computing resources for resolutions up to 1 kbp.\\	
	\textbf{Availability:} OneD is implemented as an R package available at \href{http://www.github.com/qenvio/dryhic}{http://www.github.com/qenvio/dryhic}.\\	
	\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
	\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}	online.}
\maketitle

@@

\subsection{Model}

@@

\begin{align*}
\lambda_i' &= \frac{\lambda_i}{\sum_j^n{\lambda_j}/n} \\
\hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}.
\end{align*}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
\caption{A. Non linear relationship between the number of restriction
sites per bin and the total number of contacts per bin in T47D. B. Total
number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
corrected (blue) signals are shown. The region that corresponds to the
first 20 Mbp of the chromosome is present in single copy, explaining that
the signal is approximately half compared to the rest of the chromosome.
Also notice that the raw signal is noisier than the corrected one.}
\label{fig:totals}
\end{figure}

@@

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mb in what follows). The second is
to compare the observed over expected contacts via the Pearson correlation
up to a given distance range. Compared to the first, this metric gives
more weight to changes occuring away from the diagonal. The third option
is to compute a correlation per distance stratum and then obtain a
stratum-adjusted correlation coefficient (SCC) as defined in
\cite{yang2017hicrep}.  Finally, \cite{yan2017hicspector} proposed a
reproducibility score based on the correlation between the eigenvectors of
the Laplacian matrix associated with the smallest eigenvalues. This
approach borrows the concepts of spectral clustering
\citep{von2007tutorial} and amounts to comparing high level features of
the matrix.

@@

\subsection{Copy number}

In aneuploid samples, the local coverage of the Hi-C experiment is
proportional to the number of copies of the region (a duplicated region in
a diploid genome will have on average a 50\% increase in the number of
contacts). See Figure \ref{fig:copy_number}B for a typical example.
Given that our model is based on the total number of contacts, which in
turn is related to the number of copies, we reasoned that a
preliminary test would be to check if the corrected number of contacts per
bin ($\hat{t}_i = \sum_{i}{\hat{x}_{ij}}$) reflects the copy number (as
measured by an independent technique such as array-based copy-number
segmentation) better than the alternatives.

Figure \ref{fig:totals}B shows the total number of corrected contacts
along chromosme 17 of a T47D sample. Our correction approach presents a
more stable and precise estimation of the copy number. Quantitatively
(Figure \ref{fig:copy_number}), Pearson correlation with independent copy
number data is high for both T47D and K562 cell lines (averages of 0.87
and 0.78 respectively), not only compared with raw data (averages of 0.74
and 0.63 respectively) but also with any of the alternative methods
(averages ranging 0.39-0.68 and 0.30-0.53 respectively). Similar results
were obtained using Spearman correlation (Supplementary Figure 2).

>>>>

\title[OneD]{OneD: increasing reproducibility of Hi-C Samples with
abnormal karyotypes}

@@

\abstract{
  \textbf{Motivation:} The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented advance in our understanding of genome
structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but none of them addresses the common issue of
abnormal karyotypes. Many experiments are performed in cancer cell lines,
which typically harbor large-scale copy number variations that create
visible defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.\\
	\textbf{Results:}
We observed that current normalization methods perform badly in the
presence of large-scale copy number variations, obscuring biological
viarations and enhancing batch effects. To address this issue, we
developed an alternative appoach designed to take into account chromosomal
abnormalities. The method, called OneD, increases reproducibility among
replicates of Hi-C samples with abnormal karyotype, thus significantly
outperforming previous methods.  In the absence of chromosomal
aberrantions, OneD fared equally well as state-of-the-art methods, making
it a safe choice for Hi-C normalization. OneD is fast and scales well in
terms of computing resources for resolutions up to 1 kbp.\\	
\textbf{Availability:} OneD is implemented as an R package available at
\href{http://www.github.com/qenvio/dryhic}
{http://www.github.com/qenvio/dryhic}.\\	
textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
textbf{Supplementary information:} Supplementary data are available at
\textit{Bioinformatics}	online.}

\maketitle

@@

\subsection{Model}
\label{sec:model}

@@

\begin{align}
\notag
\lambda_i' &= \frac{\lambda_i}{\sum_j^n{\lambda_j}/n} \\
\label{eq:xhat}
\hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}.
\end{align}

@@

@@

\begin{table*}
\processtable{Hi-C data sets used in this study.
\label{tab:samples}}
{\begin{tabular}{ccccccc}
  \toprule
  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set} &
  \textbf{Source} \\
  \midrule
dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D & NA \\
dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
  DpnII & CRG & T47D, K562 & NA \\
HindIII\_T0 & T47D & dilution Hi-C &
  HindIII & CRG & T47D & SRR1054341 \\
NcoI\_T0    & T47D & dilution Hi-C &
  NcoI      & CRG & T47D & SRR1054343 \\
ENCLB758KFU & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB758KFU \\
ENCLB183QHG & T47D & dilution Hi-C &
  HindIII   & UMass & T47D & ENCLB183QHG \\
HIC069  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658693 \\
HIC070  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & T47D, K562 &    SRR1658694 \\
HIC071  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
HIC074  & K562 & \textit{in situ} Hi-C &
  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
b7fa2d8db\_bfac48760 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_7bf1bf374 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_7284b867a & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_38bfd1b33 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_58e812fc2 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_c990a254e & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
b7fa2d8db\_73f11d923 & B-cell  & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
fc3e8b36a\_4bf044f18 & ES-cell & \textit{in situ} Hi-C &
  DpnII & CRG & mm10 & GSE96611 \\
GSM987817 & B-cell  & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR543428-SRR543431 \\
GSM987818 & B-cell  & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR543432-SRR543442 \\
GSM862720 & ES-cell & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR443883-SRR443885 \\
GSM862721 & ES-cell & dilution Hi-C &
  HindIII & UCSC & mm10 & SRR400251-SRR400256 \\
GSM862722 & ES-cell & dilution Hi-C &
  NcoI    & UCSC & mm10 & SRR443886-SRR443888 \\
  \botrule
\end{tabular}}{}
\end{table*}

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mb in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occuring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined in
\cite{yang2017hicrep}. Finally, the last option, proposed by
\cite{yan2017hicspector} is to measure the Pearson correlation between the
last eigenvectors of the Laplacian of the Hi-C matrix. This approach
borrows the concepts of spectral clustering \citep{von2007tutorial} and
amounts to comparing high level features of the matrix.

@@

\subsection{Bias correction}

The principle of OneD is to model explicit Hi-C biases onto a single
variable: the total amount of contacts for each bin of the matrix (see
section~\ref{sec:model}). The reason for this choice of that the total
amount of contacts is approximately proportional to the local copy number.
For instance, a duplicated region in a diploid genome will show on average
a 50\% increase in the number of contacts.

\begin{figure}[!tp]
\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
\caption{A. Non linear relationship between the number of restriction
sites per bin and the total number of contacts per bin in T47D. B. Total
number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
corrected (blue) signals are shown. The region that corresponds to the
first 20 Mbp of the chromosome is present in single copy, explaining that
the signal is approximately half compared to the rest of the chromosome.
Also notice that the raw signal is noisier than the corrected one.}
\label{fig:totals}
\end{figure}

Other biases will also affect the total amount of contacts, though not
proportionately. Figure~\ref{fig:totals}A shows the relationship
between the amount of contacts and the availability of retriction enzyme
sites. Four clouds are visible. Each corresponds to a copy number
between one and four. In all of them, the relationship flattens as the
number of restriction sites increases. OneD thus fits a non-linear
relationship between the total amount of contacts and the predictors of
bias (by default the G+C content, the number of restriction sites and the
mappability of the reads).

The biases are estimated for each bin of the genome, and each cell of the
matrix is corrected using equation (\ref{eq:xhat}). The corrected amount
of contacts is still proportional to the copy number, but it is now less
noisy. The correction thus represents a more stable and precise estimation
of the copy number. Figure \ref{fig:totals}B shows the corrected number of
contacts along chromosme 17 of a T47D sample. The corrected contacts have
a narrower range than the raw contacts.

We tested the validity of this approach against the Catalogue Of Somatic
Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure
\ref{fig:copy_number} shows the Pearson correlation between the corrected
number of contacts and the copy number estimation for both T47D and K562
cell lines. OneD achieves the highest correlations in both cases.
Actually, it is the only method that increases the agreement with the
external data. The other methods, including caICB, only make the
normalized data more divergent than the raw counts.  Similar results were
obtained using Spearman correlation (Supplementary Figure 2).

Taken together, these results indicate that OneD is an efficient way to
estimate the local copy number in order to correct for it.
